Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C

### 2024-04-20 submission/manuscript.Rmd
draft of figure 3 results

#function to get the GO ID
getGO <- function(x, df, column){
	df_test <- df
	out <- df_test[df_test$Description == x,][column][,1]
	return(out)
}

@@

The testis contains both germ cells (e.g. spermatogonia) and somatic cells (e.g. Leydig cells) that support hormone production and germline functions. We then wondered if the high enrichment of testis genes in the _Kdm5c_-KO brain reflected a failure to demarcate between the soma and germline. Gene ontology analysis of testis-biased DEGs revealed enrichment of germline-relevant ontologies, including spermatid development (`r getGO("spermatid development", testisGO, 'ID')`, q = `r signif(getGO("spermatid development", testisGO, 'p.adjust'), 3)`) and sperm axoneme assembly (`r getGO("sperm axoneme assembly", testisGO, 'ID')`, q = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust'), 3)`) (Figure 2A).

@@

We then aimed to globally characterize germline gene misexpression in _Kdm5c_-KO mice, but lacked a comprehensive list of mouse germline-enriched genes. To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice. We assessed gene expression in male and female mice at embryonic day 12, 14, and 16 embryos, as well as male postnatal day 6 and adult testes.

We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes. To more comprehensively assess the impact of sex on germline gene misexpression, we additionally categorized if germline-enriched genes displayed biased expression in the wild-type female or male germline. We defined sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes, which is consistent with the testis overall having a more unique transcriptome than the ovary[@liComprehensiveMouseTranscriptomic2017].

## _Kdm5c_-KO cells express master regulators of germline identity in an _in vitro_ model of early embryogenesis

Misexpression of germline genes in the _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity. Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the body's somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top). Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. Therefore, to elucidate the mechanism underlying aberrant germline genes expression in the mature _Kdm5c_-KO brain, we first assessed their expression in male _Kdm5c_-KO EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022].

* We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom). Additionally, _Kdm5c_-KO EpiLCs downregulated ESC pluripotency genes and upregulated EpiLC differentiation genes similar to wild-type (Figure 3B).

* We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and observed widespread dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis (Figure 3C).

* Of these testis genes, we also observed misexpression of germline-enriched genes, including _Cyct_ (Figure 3D)

* To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. The majority of EpiLC germline DEGs were unique to EpiLCs, with only a few shared across brain sequencing datasets (Figure 3E). Only two germline genes were consistently misexpressed in all datasets, _DNA segment, Chr 1, Pasteur Institute 1 (D1Pas1)_ and _Testis expressed gene 14 (Tex14)_.

* We then compared the typical function of brain and EpiLC germline DEGs and found EpiLC germline DEGs displayed strong enrichment of meiosis-related gene ontologies, including XXXXX (Figure 3F).

* Surprisingly, germline DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but they are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].

* _Dazl_ is a translational regulator essential for germ cell function[@liDAZLMasterTranslational2019]. Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (XY _Kdm5c_-KO EpiLCs p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_ mutant cellular function.

* Therefore, we then stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H). We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test, XY KO versus WT), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020].

>>>>

#function to get the GO ID
getGO <- function(x, df, column, cluster){
	df_test <- df

	if (cluster == "no") {
		out <- df_test[df_test$Description == x,][column][,1]
	} else {
		out <- df_test[df_test$Description == x & df_test$Cluster == cluster,][column][,1]
	}
	return(out)
}

@@

The testis contains both germ cells (e.g. spermatogonia) and somatic cells (e.g. Leydig cells) that support hormone production and germline functions. We then wondered if the high enrichment of testis genes in the _Kdm5c_-KO brain reflected a failure to demarcate between the soma and germline. Gene ontology analysis of testis-biased DEGs revealed enrichment of germline-relevant ontologies, including spermatid development (`r getGO("spermatid development", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (`r getGO("sperm axoneme assembly", testisGO, 'ID', "no")`, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).

@@

We then aimed to globally characterize germline gene misexpression in _Kdm5c_-KO mice, but lacked a comprehensive list of mouse germline-enriched genes. To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice. We assessed gene expression in male and female mice at embryonic day 12, 14, and 16 embryos, as well as male postnatal day 6 and adult testes. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes.

## _Kdm5c_-KO epiblast-like cells aberrantly express master regulators of germline identity

```{r, epilcs, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}

#testis gene ontology results
epigerm <- read.csv("../results/DESeq2/germDEGs/germDEGs_EpiLC_XY5cKO.csv")
GO_EpiBrain <- read.csv("../results/GO_EpiLC_vs_Brain.csv")

epi_restab <- read.csv("../data/processed/restable_EpiLC_XY5cKO.csv", row.names = 1)
epi_restab$ENSEMBL <- rownames(epi_restab)

zscan4d <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000090714',]['padj'][,1]

Misexpression of germline genes in the _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity. Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the body's somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top). Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. Therefore, we assessed if KDM5C was necessary for initial germline gene silencing by assessing their expression in male _Kdm5c_-KO EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022].

We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom) and found no significant difference in primed pluripotency genes that indicate proper EpiLC differentiation (Figure 3B). We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and again observed widespread dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3C).

Using our curated list, we found `r nrow(epigerm)` germline-enriched genes were misexpressed in _Kdm5c_-KO EpiLCs, including _Cyct_ (Figure 3D). To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across sequencing datasets (Figure 3E).

We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`).Surprisingly, germline DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo. However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`).

_Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019]. Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_ mutant cellular function. We therefore stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H). We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including master regulators of germline identity that can be translated into protein.
